---
title: "Eksamensæt"
output:
  word_document: default
  pdf_document: default
date: "2022-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use data in constant prices (preferable).
- Real GDP
- Real house prices
- Real stock of credit
- Stock prices

Make sure that you first remove the seasonality (run an F-test to see
if this is necessary). Make sure to write the regression equations and
hypotheses when you perform these procedures.


A time series is a collection of random variables ordered in time. time series helps us to describe and illustrate inferences concerning the development of the data. 

Decomposing a time series into components:
$$X_t=seasonal_t+trend_t+irregular_t$$
seasonallity can be shown at regular intervals (monthly or quarterly)
the trend shows the trend of the variables
irregular component even though lacks a well-defined pattern is
somewhat predictable

When seasonal fluctuations and trend are removed from an
economic time series, the remaining irregular component in the
medium to long term exhibits a cyclical pattern (due to expansions and
contractions in the economy)


When seasonal fluctuations and trend are removed from an economic time series,
the remaining irregular component in the medium to long term exhibits a cyclical pattern (due to expansions and contractions in the economy)
The "cyclical pattern" of the irregular component is of great interest to economists

The difference between a cyclical and a seasonal component is that:
1. Seasonality occurs at regular (seasonal) intervals (e.g., monthly or quarterly in typical macro data)
2. Cyclical factors have usually a longer duration that varies from cycle to cycle

if we look at various statistical techniques that can be used to
decompose a time series into different components

isolating seasonality, isolating trend and the remainder is irregular fluctuation$$Y_t=α + β_2D_2 + β_3D_3 + β_4D_4 + ε_t$$
We can now run the F-test for the nulhypothesis $H_0:β_2 = β_3 = β_4 = 0$

$$X_t-seasonal_t-trend_t=irregular_t$$

we are now going to isolate seasonality by using the dummy variable method

$$Y_t=α + β_2D_2 + β_3D_3 + β_4D_4 + ε_t$$
We can now run the F-test for the nulhypothesis $H_0:β_2 = β_3 = β_4 = 0$

if we want to reject the $H_0$ we want $Y_t$ to display a significant seasonal pattern

$$\hat{\epsilon_t} =(Y_t − \hat{Y_t})$$
is our seasonally adjusted time series


## Loading data
```{r}
# Load necessary libraries
library(dplyr)
library(readxl)
library(zoo)


DanishBNP <- read_excel("Data/danishBNP.xlsx", sheet = 1)
colnames(DanishBNP) <- c("Year_K", "BNP")
DanishBNP$Year_K <- gsub("K", " Q", DanishBNP$Year_K)
DanishBNP$YearQtr <- as.yearqtr(DanishBNP$Year_K, format = "%Y Q%q")

DanishInflation <- read_excel("Data/danishinflation.xlsx", sheet = 1)
DanishInflation <- DanishInflation[3:nrow(DanishInflation), ]
colnames(DanishInflation) <- c("year", "Inflation")
DanishInflation <- na.omit(DanishInflation)


DanishInterestReate <- read_excel("Data/danishinterestrate.xlsx", sheet = 1)
DanishInterestReate <- DanishInterestReate[3:37,]
colnames(DanishInterestReate) <- c("year", "InterestRate")
DanishInterestReate <- na.omit(DanishInterestReate)


DanishUnemployment <- read_excel("Data/danishunemployment.xlsx", sheet = 1)


SWEBNP <- read_excel("Data/SwedenGDP.xlsx", sheet = 1)
colnames(SWEBNP) <- c("Year_K", "BNP")
SWEBNP$Year_K <- gsub("K", " Q", SWEBNP$Year_K)
SWEBNP$YearQtr <- as.yearqtr(SWEBNP$Year_K, format = "%Y Q%q")


SWEInflation <- read_excel("Data/sweden-inflation.xlsx", sheet = 1)
colnames(SWEInflation) <- c("Year", "Inflation")


SWEInterestReate <- read_excel("Data/swedeninterestrate.xls", sheet = 1)
colnames(SWEInterestReate) <- c("Year", "Interest_Reate")


SWEUnemployment <- read_excel("Data/sweden-unemployment.xls", sheet = 1)
SWEUnemployment <- SWEUnemployment[-1, ]
colnames(SWEUnemployment) <- c("Year", "Unemployment_Rate")
SWEUnemployment$Year <- as.Date(as.numeric(SWEUnemployment$Year), origin = "1900-01-01")
SWEUnemployment$Year <- format(SWEUnemployment$Year, "%Y")
SWEUnemployment$Year <- as.numeric(SWEUnemployment$Year)

OMXC20 <- read.csv("Data/OMXC20.csv")
OMXS30 <- read.csv("Data/OMXS30.csv")


# Ensure numeric columns are indeed numeric
DanishInflation$Inflation <- as.numeric(DanishInflation$Inflation)
DanishInterestReate$InterestRate <- as.numeric(DanishInterestReate$InterestRate)
DanishUnemployment$`Procent af arbejdsstyrken` <- as.numeric(DanishUnemployment$`Procent af arbejdsstyrken`)

SWEInflation$Inflation <- as.numeric(SWEInflation$Inflation)
SWEInterestReate$Interest_Reate <- as.numeric(SWEInterestReate$Interest_Reate)
SWEUnemployment$Unemployment_Rate <- as.numeric(SWEUnemployment$Unemployment_Rate)

OMXC20$Close.Last <- as.numeric(OMXC20$Close.Last)
OMXS30$Close.Last <- as.numeric(OMXS30$Close.Last)

```

## Plot OMXC20 vs OMXS30
```{r}
# Load necessary libraries
library(dplyr)
library(zoo)

# Convert Date columns to Date format
OMXC20$Date <- as.Date(OMXC20$Date, format = "%m/%d/%Y")
OMXS30$Date <- as.Date(OMXS30$Date, format = "%m/%d/%Y")

# Ensure Close.Last is numeric
OMXC20$Close.Last <- as.numeric(OMXC20$Close.Last)
OMXS30$Close.Last <- as.numeric(OMXS30$Close.Last)

# Remove rows with NA values
OMXC20 <- na.omit(OMXC20)
OMXS30 <- na.omit(OMXS30)

# Select necessary columns and rename Close.Last
OMXC20_subset <- OMXC20 %>%
  select(Date, Close.Last) %>%
  rename(OMXC20_Close = Close.Last)

OMXS30_subset <- OMXS30 %>%
  select(Date, Close.Last) %>%
  rename(OMXS30_Close = Close.Last)

# Merge datasets using full_join to include all dates
merged_data <- full_join(OMXC20_subset, OMXS30_subset, by = "Date")

# Sort by Date
merged_data <- merged_data %>% arrange(Date)

# Forward fill missing values
merged_data$OMXC20_Close <- na.locf(merged_data$OMXC20_Close, na.rm = FALSE)
merged_data$OMXS30_Close <- na.locf(merged_data$OMXS30_Close, na.rm = FALSE)

# Remove any rows where either Close price is still NA
merged_data <- merged_data[!is.na(merged_data$OMXC20_Close) & !is.na(merged_data$OMXS30_Close), ]

# Normalize starting from the first date where both indices have data
merged_data$OMXC20_Normalized <- merged_data$OMXC20_Close / merged_data$OMXC20_Close[1]
merged_data$OMXS30_Normalized <- merged_data$OMXS30_Close / merged_data$OMXS30_Close[1]

# Plot the normalized data
plot(merged_data$Date, merged_data$OMXC20_Normalized, type = "l", col = "blue",
     main = "Relative Close.Last Over Time for Scandinavian Indices",
     xlab = "Date", ylab = "Relative Close.Last (Start = 1)")

lines(merged_data$Date, merged_data$OMXS30_Normalized, col = "green")

legend("topright", legend = c("OMXC20", "OMXS30"),
       col = c("blue", "green"), lty = 1)

```
## ADF Test

```{r}
library(tseries)

# Run the ADF test on market Close prices
adf_test_OMXC20 <- adf.test(OMXC20$Close.Last, alternative = "stationary")
adf_test_OMXS30 <- adf.test(OMXS30$Close.Last, alternative = "stationary")

#Denmark
adf_test_DBNP <- adf.test(DanishBNP$BNP, alternative = "stationary")
adf_test_Dinflation <- adf.test(DanishInflation$Inflation, alternative = "stationary")
adf_test_Dinterest <- adf.test(DanishInterestReate$InterestRate, alternative = "stationary")
adf_test_Dunemployment <- adf.test(DanishUnemployment$`Procent af arbejdsstyrken`, alternative = "stationary")

#Sweden
adf_test_SBNP <- adf.test(SWEBNP$BNP, alternative = "stationary")
adf_test_Sinflation <- adf.test(SWEInflation$Inflation, alternative = "stationary")
adf_test_Sinterest <- adf.test(SWEInterestReate$Interest_Reate, alternative = "stationary")
adf_test_Sunemployment <- adf.test(SWEUnemployment$Unemployment_Rate, alternative = "stationary")


# Display the results
cat("ADF Test Result for OMXC20 Close Prices:\n")
print(adf_test_OMXC20)
cat("\nADF Test Result for OMXS30 Close Prices:\n")
print(adf_test_OMXS30)

print(adf_test_DBNP)
print(adf_test_Dinflation)
print(adf_test_Dinterest)
print(adf_test_Dunemployment)

print(adf_test_SBNP)
print(adf_test_Sinflation)
print(adf_test_Sinterest)
print(adf_test_Sunemployment)

```
## first-order differencing

```{r}
# Load necessary libraries
library(dplyr)
library(readxl)
library(zoo)
library(tseries)
library(ggplot2)



## First-Order Differencing

# For Denmark
diff_DBNP <- diff(DanishBNP$BNP)
diff_Dinflation <- diff(DanishInflation$Inflation)
diff_Dinterest <- diff(DanishInterestReate$InterestRate)
diff_Dunemployment <- diff(DanishUnemployment$`Procent af arbejdsstyrken`)

# For Sweden
diff_SBNP <- diff(SWEBNP$BNP)
diff_Sinflation <- diff(SWEInflation$Inflation)
diff_Sinterest <- diff(SWEInterestReate$Interest_Reate)
diff_Sunemployment <- diff(SWEUnemployment$Unemployment_Rate)

# For OMXC20 and OMXS30
diff_OMXC20 <- diff(OMXC20$Close.Last)
diff_OMXS30 <- diff(OMXS30$Close.Last)

## Apply the ADF Test on Differenced Series

# Function to display ADF test results
display_adf_result <- function(adf_result, variable_name) {
  cat("\nADF Test Result for", variable_name, ":\n")
  print(adf_result)
}

# Denmark
adf_diff_DBNP <- adf.test(diff_DBNP, alternative = "stationary")
display_adf_result(adf_diff_DBNP, "Differenced Danish GDP (BNP)")

adf_diff_Dinflation <- adf.test(diff_Dinflation, alternative = "stationary")
display_adf_result(adf_diff_Dinflation, "Differenced Danish Inflation")

adf_diff_Dinterest <- adf.test(diff_Dinterest, alternative = "stationary")
display_adf_result(adf_diff_Dinterest, "Differenced Danish Interest Rate")

adf_diff_Dunemployment <- adf.test(diff_Dunemployment, alternative = "stationary")
display_adf_result(adf_diff_Dunemployment, "Differenced Danish Unemployment")

# Sweden
adf_diff_SBNP <- adf.test(diff_SBNP, alternative = "stationary")
display_adf_result(adf_diff_SBNP, "Differenced Swedish GDP (BNP)")

adf_diff_Sinflation <- adf.test(diff_Sinflation, alternative = "stationary")
display_adf_result(adf_diff_Sinflation, "Differenced Swedish Inflation")

adf_diff_Sinterest <- adf.test(diff_Sinterest, alternative = "stationary")
display_adf_result(adf_diff_Sinterest, "Differenced Swedish Interest Rate")

adf_diff_Sunemployment <- adf.test(diff_Sunemployment, alternative = "stationary")
display_adf_result(adf_diff_Sunemployment, "Differenced Swedish Unemployment")

# OMXC20 and OMXS30
adf_diff_OMXC20 <- adf.test(diff_OMXC20, alternative = "stationary")
display_adf_result(adf_diff_OMXC20, "Differenced OMXC20 Close Prices")

adf_diff_OMXS30 <- adf.test(diff_OMXS30, alternative = "stationary")
display_adf_result(adf_diff_OMXS30, "Differenced OMXS30 Close Prices")


```

## 2nd & 3rd order diff

```{r}
diff2_Dinterest <- diff(DanishInterestReate$InterestRate, differences = 2)
diff2_Sunemployment <- diff(SWEUnemployment$Unemployment_Rate, differences = 2)

adf_diff2_Dinterest <- adf.test(diff2_Dinterest, alternative = "stationary")
display_adf_result(adf_diff2_Dinterest, "2nd order Differenced DanishInterestRate")

adf_diff2_Sunemployment <- adf.test(diff2_Sunemployment, alternative = "stationary")
display_adf_result(adf_diff2_Sunemployment, "2nd order Differenced Swedish unemployment")


```
```{r}
diff3_Sunemployment <- diff(SWEUnemployment$Unemployment_Rate, differences = 3)
adf_diff3_Sunemployment <- adf.test(diff3_Sunemployment, alternative = "stationary")
display_adf_result(adf_diff3_Sunemployment, "3nd order Differenced Swedish unemployment")
```
# plots

```{r}
library(ggplot2)
library(gridExtra)

# GDP Plot
p1 <- ggplot(SwedenData, aes(x = Date, y = GDP)) +
  geom_line(color = "blue") +
  labs(title = "Sweden GDP", x = "Year", y = "GDP") +
  theme_minimal()

# Inflation Plot
p2 <- ggplot(SwedenData, aes(x = Date, y = Inflation)) +
  geom_line(color = "red") +
  labs(title = "Sweden Inflation", x = "Year", y = "Inflation Rate (%)") +
  theme_minimal()

# Unemployment Plot
p3 <- ggplot(SwedenData, aes(x = Date, y = Unemployment)) +
  geom_line(color = "green") +
  labs(title = "Sweden Unemployment", x = "Year", y = "Unemployment Rate (%)") +
  theme_minimal()

# Interest Rate Plot
p4 <- ggplot(SwedenData, aes(x = Date, y = InterestRate)) +
  geom_line(color = "purple") +
  labs(title = "Sweden Interest Rate", x = "Year", y = "Interest Rate (%)") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)



```



```{r}
library(readxl)
data <- read_excel("C:/Users/celin/OneDrive/Skrivebord/Cand.oecon/5. Semester - Oecon BA/Økonometri/Assignments/data.xlsx")


#Summarize the data:
summary(data)

#Vi opsætter nu variablerne som tidsserier:
Real_credit <- ts(data$real_Credit, start = c(1950,2), end = c(2018, 2), frequency = 4)
length(Real_credit)

Real_GDP <- ts(data$real_gdp, start = c(1950,2), end = c(2018, 2), frequency = 4)
length(Real_GDP)

Real_house_prices <- ts(data$`real house prices`, start = c(1950,2), end = c(2018, 2), frequency = 4)
length(Real_house_prices)

Stock_prices <- ts(data$`stock prices`, start = c(1950,2), end = c(2018, 2), frequency = 4)
length(Stock_prices)
```

Herefter plotter vi tidsserierne

```{r}
plot(Real_credit, xlab = "Year", ylab = "Real Credit", lwd = "2")
plot(Real_GDP, xlab = "Year", ylab = "Real GDP", lwd = "2")
plot(Real_house_prices, xlab = "Year", ylab = "Real house prices", lwd = "2")
plot(Stock_prices, xlab = "Year", ylab = "Stock prices", lwd = "2")
```

Vi vil nu isolere for seasonallity i hver af variablerne

```{r}
#Tester for Real credit
Real_credit_X <- cycle(Real_credit)
Real_credit_q1 <- ifelse(Real_credit_X == "1", 2, 15)
Real_credit_q2 <- ifelse(Real_credit_X == "2", 5, 15)
Real_credit_q3 <- ifelse(Real_credit_X == "3", 8, 15)
Real_credit_q4 <- ifelse(Real_credit_X == "4", 11, 15)
summary(Real_credit_y.reg <- y.reg <- lm(Real_credit ~ Real_credit_q1 + Real_credit_q2 + Real_credit_q3))
```
Real_credit kan vi ikke afvise null hypotesen - F-statistic er usignifikant. 
ingen andre signifikante quarters - ikke tydeligt bevis for seasonality for Real Credit

```{r}
#Tester for real GDP
Real_GDP_X <- cycle(Real_GDP)
Real_GDP_q1 <- ifelse(Real_GDP_X == "1", 2, 15)
Real_GDP_q2 <- ifelse(Real_GDP_X == "2", 5, 15)
Real_GDP_q3 <- ifelse(Real_GDP_X == "3", 8, 15)
Real_GDP_q4 <- ifelse(Real_GDP_X == "4", 11, 15)
summary(Real_GDP_y.reg <- y.reg <- lm(Real_GDP ~ Real_GDP_q1 + Real_GDP_q2 + Real_GDP_q3))
```

```{r}
#Tester for Real House prices
Real_house_prices_X <- cycle(Real_house_prices)
Real_house_prices_q1 <- ifelse(Real_house_prices_X == "1", 2, 15)
Real_house_prices_q2 <- ifelse(Real_house_prices_X == "2", 5, 15)
Real_house_prices_q3 <- ifelse(Real_house_prices_X == "3", 8, 15)
Real_house_prices_q4 <- ifelse(Real_house_prices_X == "4", 11, 15)
summary(Real_house_prices_y.reg <-y.reg <- lm(Real_house_prices ~ Real_house_prices_q1 + Real_house_prices_q2 + Real_house_prices_q3))
```

```{r}
#Tester for Stock prices
Stock_prices_X <- cycle(Stock_prices)
Stock_prices_q1 <- ifelse(Stock_prices_X == "1", 2, 15)
Stock_prices_q2 <- ifelse(Stock_prices_X == "2", 5, 15)
Stock_prices_q3 <- ifelse(Stock_prices_X == "3", 8, 15)
Stock_prices_q4 <- ifelse(Stock_prices_X == "4", 11, 15)
summary(Stock_prices_y.reg <- y.reg <- lm(Stock_prices ~ Stock_prices_q1 + Stock_prices_q2 + Stock_prices_q3))
```





# Plot the series with and without seasonality on the same graph.
Vi vil starte med at undersøge $\hat{\epsilon_t}$ for tidsserierne
```{r}
res_Real_credit <- residuals(Real_credit_y.reg)
res_Real_GDP <- residuals(Real_GDP_y.reg)
res_Real_house_prices <- residuals(Real_house_prices_y.reg)
res_Stock_prices <- residuals(Stock_prices_y.reg)
```

Plotter the residuals:
```{r}
plot(ts(res_Real_credit, start = 1950, frequency = 4), lwd = 2)
plot(ts(res_Real_GDP, start = 1950, frequency = 4), lwd = 2)
plot(ts(res_Real_house_prices, start = 1950, frequency = 4), lwd = 2)
plot(ts(res_Stock_prices, start = 1950, frequency = 4), lwd = 2)
```

Plotter the fitted values:
```{r}
plot(ts(fitted.values(Real_credit_y.reg), start = 1950, frequency = 4), lwd = 2)
plot(ts(fitted.values(Real_GDP_y.reg), start = 1950, frequency = 4), lwd = 2)
plot(ts(fitted.values(Real_house_prices_y.reg), start = 1950, frequency = 4), lwd = 2)
plot(ts(fitted.values(Stock_prices_y.reg), start = 1950, frequency = 4), lwd = 2)
```

Der plottes nu de-seasonalised series i forbindelse med den oprindelige
```{r}
Real_credit_yhat <- residuals(Real_credit_y.reg) + mean(fitted.values(Real_credit_y.reg))
ts.plot(Real_credit, Real_credit_yhat, lty = c(3, 1), lwd = c(4, 1))
legend(1950, max(Real_credit), c("Real_credit", "Seasonally adjusted Real credit (Real_credit_yhat)"), lty = c(3,
1))

Real_GDP_yhat <- residuals(Real_GDP_y.reg) + mean(fitted.values(Real_GDP_y.reg))
ts.plot(Real_GDP, Real_GDP_yhat, lty = c(3, 1), lwd = c(4, 1))
legend(1950, max(Real_GDP), c("Real_GDP", "Seasonally adjusted Real GDP (Real_GDP_yhat)"), lty = c(3,
1))


Real_house_prices_yhat <- residuals(Real_house_prices_y.reg) + mean(fitted.values(Real_house_prices_y.reg))
ts.plot(Real_house_prices, Real_house_prices_yhat, lty = c(3, 1), lwd = c(4, 1))
legend(1950, max(Real_house_prices), c("Real_house_prices", "Seasonally adjusted Real house prices (Real_house_prices_yhat)"), lty = c(3,
1))

Stock_prices_yhat <- residuals(Stock_prices_y.reg) + mean(fitted.values(Stock_prices_y.reg))
ts.plot(Stock_prices, Stock_prices_yhat, lty = c(3, 1), lwd = c(4, 1))
legend(1950, max(Stock_prices), c("Stock_prices", "Seasonally adjusted Stock prices (Stock_prices_yhat)"), lty = c(3,
1))
```




# If there is a trend in the data, detrend the variables using HP-filter. Use the HP-filter with λ = 1600. Don’t forget to re-normalize the series

When isolating the trend in a lineat time trends we can run the following regression equation

$$output_t= α + βyear_t + ε_t$$
If there is a linear time trend, $β$ will have a significant t-value.
If the t-value is significant, then we use $\hat{ε_t}$ as our variable for
detrended output.
If the t-value is not significant, then we continue the exercise using
$output_t$

if the time trend is non linear it will se necessary to remove a quadratic time trend



We will use the Hodrick-Prescott Filter

The HP gilter allows for a more flexible trend.
$X_t$ is the sum os a cyclical componenet, $X_t^c$
a trend component $X_t^t$
a contemporaneous shock $ε_t$

$$X_t=X^c_t + X^t_t + ε_t$$
a parameter which reflects the relative variance of the trend component to the cyclical component, $λ$

a given $λ$, choose the trend component, $X^t_t$, which minimizes the following loss function


Modellerne estimeres:
```{r}
date <- data$date
summary(Real_credit_trend_y.reg <- y.reg <- lm(Real_credit ~ date))
```

```{r}
summary(Real_GDP_trend_y.reg <- y.reg <- lm(Real_GDP ~ date))
```

```{r}
summary(Real_house_prices_trend_y.reg <- y.reg <- lm(Real_house_prices ~ date))
```

```{r}
summary(Stock_prices_trend_y.reg <- y.reg <- lm(Stock_prices ~ date))
```

Her ses det at alle t-value er signifikante derfor ses på $\hat{\epsilon_t}$
```{r}
ty_RC = residuals(Real_credit_trend_y.reg)
ty1_RC = ty_RC + Real_credit[1]

ty_RGDP = residuals(Real_GDP_trend_y.reg)
ty1_RGDP = ty_RGDP + Real_GDP[1]

ty_RHP = residuals(Real_house_prices_trend_y.reg)
ty1_RHP = ty_RHP + Real_house_prices[1]

ty_SP = residuals(Stock_prices_trend_y.reg)
ty1_SP = ty_SP + Stock_prices[1]
```

Undersøg $\hat{Output_t}$. Dette er din tidstrend, det vil sige på lang sigt
(lineær) væksttendens for produktionen. Den kan fås i R ved skrivning:
```{r}
trend_RC <- fitted.values(Real_credit_trend_y.reg)
trend_RGDP <- fitted.values(Real_GDP_trend_y.reg)
trend_RHP <- fitted.values(Real_house_prices_trend_y.reg)
trend_SP <- fitted.values(Stock_prices_trend_y.reg)
```


Tegn en graf af den originale serie, sammen med tidens trend og residuals:
```{r}
ts.plot(Real_credit, trend_RC, ty1_RC, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Real_credit), c("Real credit (original series)", "trend", "Real credit (detrended series)"), lty = (1:3))

ts.plot(Real_GDP, trend_RGDP, ty1_RGDP, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Real_GDP), c("Real GDP (original series)", "trend", "Real GDP (detrended series)"), lty = (1:3))

ts.plot(Real_house_prices, trend_RHP, ty1_RHP, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Real_house_prices), c("Real house prices (original series)", "trend", "Real house prices (detrended series)"), lty = (1:3))

ts.plot(Stock_prices, trend_SP, ty1_SP, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Stock_prices), c("Stock prices (original series)", "trend", "Stock prices (detrended series)"), lty = (1:3))
```
BNP - Andre faktore der påvirker bnp
hvis man fjerne seasonality og trend kan dataen blive stationary, men det gælder ikke alle

Som det kan ses er der trend i serien og først er den fundet som liniær trend nu findes den som mere volatil trend ved brug af HP filter
```{r}
library(mFilter)
lambda <- 1600
hp_RC <- hpfilter(Real_credit, lambda)
hptrend_RC <- hp_RC$trend

library(mFilter)
lambda <- 1600
hp_RGDP <- hpfilter(Real_GDP, lambda)
hptrend_RGDP <- hp_RGDP$trend

library(mFilter)
lambda <- 1600
hp_RHP <- hpfilter(Real_house_prices, lambda)
hptrend_RHP <- hp_RHP$trend

library(mFilter)
lambda <- 1600
hp_SP <- hpfilter(Stock_prices, lambda)
hptrend_SP <- hp_SP$trend
```

En H-P detrended serie kan oprettes i R ved at skrive:
```{r}
hp_dRC <- Real_credit - hptrend_RC
hpy1_dRC <- hp_dRC + Real_credit[1]

hp_dRGDP <- Real_GDP - hptrend_RGDP
hpy1_dRGDP <- hp_dRGDP + Real_GDP[1]

hp_dRHP <- Real_house_prices - hptrend_RHP
hpy1_dRHP <- hp_dRHP + Real_house_prices[1]

hp_dSP <- Stock_prices - hptrend_SP
hpy1_dSP <- hp_dSP + Stock_prices[1]
```

Den nye serie plottet
```{r}
ts.plot(Real_credit, hptrend_RC, hpy1_dRC, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Real_credit), c("Real credit (original series)", "hptrend", "Real credit (detrended series)"), lty = (1:3))

ts.plot(Real_GDP, hptrend_RGDP, hpy1_dRGDP, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Real_GDP), c("Real GDP (original series)", "hptrend", "Real GDP (detrended series)"), lty = (1:3))

ts.plot(Real_house_prices, hptrend_RHP, hpy1_dRHP, lty = c(1,3), lwd = c(2, 2, 2), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Real_house_prices), c("Real house prices (original series)", "hptrend", "Real house prices (detrended series)"), lty = (1:3))

ts.plot(Stock_prices, hptrend_SP, hpy1_dSP, lty = c(1,3), lwd = c(3, 3, 3), xlab = "Year",
col = c("black", "red", "blue"))
legend(1950, max(Stock_prices), c("Stock prices (original series)", "hptrend", "Stock prices (detrended series)"), lty = (1:3))
```

Der henvises også til forskellen mellem faktisk produktion og trend
som "output gap" (blå linje i ovenstående plot). Trend i så fald
repræsenterer "potentielt output"




# Explore the dynamic correlations (or cross-correlations) using real GDP as a reference variable. You can make a Table like the one in Lecture no. 2. Also, show the plots of dynamic correlations. The plots should clearly show all the statistically relevant lags 

```{r}
RGDP <- as.numeric(Real_GDP)
RC <- as.numeric(Real_credit)
RHP <- as.numeric(Real_house_prices)
SP <- as.numeric(Stock_prices)

cc_RGDP_RC <- ccf(RGDP, RC, 1, pl = F)
cc_RGDP_RC
cc_RGDP_RHP <- ccf(RGDP, RHP, 1, pl = F)
cc_RGDP_RHP
cc_RGDP_SP <- ccf(RGDP,SP, 1, pl = F)
cc_RGDP_SP
```
Plots:
```{r}
plot(cc_RGDP_RC, ci=0.95, ylab="", xlab="", main="") 
plot(cc_RGDP_RHP, ci=0.95, ylab="", xlab="", main="")
plot(cc_RGDP_SP, ci=0.95, ylab="", xlab="", main="") 
```



Contemporaneous correlation: It is the correlation between two
variables in time period (t)
To calculate the contemporaneous correlation between x(t) and y(t)
write: cor(x,y)
To calculate a series of dynamic or cross-correlations between x(t)
and lags and leads of y(t) we use the cross-correlation function ccf


Note: reference variable is y(t) - you can see how lags and leads of x are correlated
with y(t).
The dotted blue line shows 95 percent confidence intervals - in this case we can
see that all correlations are statistically significant. If you write ci=99 in the code,
it will set ci to 99 percent.



# Calculate cyclical volatility of your data.
Vi opstiller formlen for at udregne cyclical volatility:
$$
\text{Volatility x} = s.d. [(\frac{x_t-trend(x)_t}{trend(x)_t})*100] 
$$

```{r}
library(mFilter)
lambda <- 1600
hp = hpfilter(RC, lambda)
trend_credit <- hp$trend

credit_dt <- Real_credit - trend_credit

v_credit <- ((Real_credit - trend_credit) * 100/trend_credit)

sd(v_credit)


hp = hpfilter(Real_GDP, lambda)
trend_gdp <- hp$trend

gdp_dt <- Real_GDP - trend_gdp

v_gdp <- ((Real_GDP - trend_gdp) * 100/trend_gdp)

sd(v_gdp)


hp = hpfilter(Real_house_prices, lambda)
trend_HP <- hp$trend

HP_dt <- Real_house_prices - trend_HP

v_HP <- ((Real_house_prices - trend_HP) * 100/trend_HP)

sd(v_HP)


hp = hpfilter(Stock_prices, lambda)
trend_SP <- hp$trend

SP_dt <- Stock_prices - trend_SP

v_SP <- ((Stock_prices - trend_SP) * 100/trend_SP)

sd(v_SP)
```

Herefter opsættes dette graphisk
```{r}
plot(date,credit_dt, t="l") 

plot(date, gdp_dt, t="l")

plot(date, HP_dt, t="l")

plot(date, SP_dt, t="l")
```


# Do a short and precise commentary on the correlations and volatility (e.g., which variable is the most and least volatile, which variables are procyclical and countercyclical, which variables lag and lead the business cycle, etc.).


```{r}
library(ggplot2)
library(tidyverse)

p = ggplot() + 
  geom_line(aes(x = date, y = credit_dt), color = "blue") +
  geom_line(aes(x = date, y = gdp_dt), color = "red") +
  geom_line(aes(x = date, y = HP_dt), color = "green", lwd = 2) +
  geom_line(aes(x = date, y = SP_dt), color = "black") +
  xlab('Dates') +
  ylab('')

print(p)
```

Den mest volatible variabel er stock prices, hvor real house prices, real GDP og real credit er langt mindre volatil men stadig mildt volatil i samme rækkefølge som opsat. Hvor real house prices og stock prices er countercyclical hvilket kan ses at deres udslag næsten ikke forekommer sammenlignet med de andre variable mens RGDP og Real credit er pro cyclical. Hertil real GDP og Real credit lede buissnes cycle mens stock og house prices vil lag grundet deres påvirkning af GDP og credit. 







